### 2024-07-02 latex/HP4_comm.tex
new input files

%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY? Judit: kalā is like śakti (initial source)?
%Refer to TĀK

@@

%%%%%%%%%%
\subsection*{4.0*7}
\begin{translation}[hp04_000_7]
Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation †arises† [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
\end{translation}

@@

\begin{philcomm}[hp04_000_7]
The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}. Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}. We are unable to make sense of \emph{sthiti} here, and so have cruxed this verse quarter and emended to \emph{muktiḥ sthitā} (`liberation has happened').
%?? JB. Rather than emend to muktiḥ sthitā, I think we should consider muktisthitiḥ (etaTwo and deltaOne), which is similar to jīvanmuktasthitiṃ in the Mokṣopāya, LYV, MSS, etc. We could translate it as Sternbach below ('state of liberation'). JM: I also don't think we should emend. muktisthitiḥ is possible (but MU parallel not great because of mukti vs mukta). We could also consider muktiḥ sthitiḥ “liberation [and] stability”
% Mokṣopāya
% jīvanmuktasthitiṃ brahman kṛtvā rāghavam āditaḥ|\\
% kramāt kathaya me nityaṃ bhaviṣyāmi sukhī yathā||\,1||
%  "jīvanmuktasthitiṃ" jīvanmuktamaryādāṃ
% Yogavāsiṣṭha
% vicārajā nijā jīvanmuktasthitir ihoditā |
% muninābhyāsahīnasya vyādhasya tvanavasthitiḥ || VII. 
% MSS_9744-1 
% kāryākārye kim api satataṃ naiva kartṛtvam asti  
% jīvanmuktasthitir avagato dagdhavastrāvabhāsaḥ /<BR>
% There is not always an agent in every action, good or bad ; for, the state of a jivanmukta [one who is released from bondage while living in the world] is considered to be similar to that of the splendour of a burnt cloth [which can neither be clean nor dirty]. (Sternbach's translation)
\end{philcomm}

@@

\begin{philcomm}[hp04_000_9]
In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of \emph{suṣumnā}.
\end{philcomm}
%?? MD: vilīyate or pralīyate?

@@

%%%%%%%%%%
\subsection*{4.0*10}
\begin{translation}[hp04_000_10]
For the yogi whose \emph{kuṇḍalinī} has awakened and who has given up all activity, the innate state automatically arises.
\end{translation}

@@

%?? MD: prajāyate or prakāśate?

%%%%%%%%%%
\subsection*{4.0*11}
\begin{translation}[hp04_000_11]
When the breath is flowing in the central channel and the mind enters the void, the knower of yoga uproots all [past] actions.
\end{translation}
%?? JB: all [fruits of past] actions

@@

%\begin{philcomm}[hp04_000_12]
%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
%?? JB should we include this note on gamma's omission? Is it necessary? JM: no, I've percented it

@@

%%%%%%%%%%
\subsection*{4.0*13}
\begin{translation}[hp04_000_13]
When the mind has obtained equanimity and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises even for the person whose hope is for that.
\end{translation}
% ?? JB: we didnt translate this verse in the workshop. I've adapted my translation from a similar verse in the Amaraugha. JM: equanimity is tautologous with mind, “When equanimity has been obtained"? Also I've changed tr. to reflect tadā.

\begin{sources}[hp04_000_13]
\emph{Amaraugha} 7
\begin{versinnote}
\tl{citte tu sattvam āpanne vāyau vrajati madhyame |\\+}
\tl{tadāmaraughavajrolis tadāśājīvitasya ca ||\\+}
\tl{\var{citte tu sattvam ] $\beta$ : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
tadāmaraughavajrolis ] \textit{conj.} :
tadāmaraughavajroli Ga :
tadāmaraudhavajrolī Ae :
eṣāmarauli vajroli Gb :
eṣāmarāli vajroli Ba :
eṣāmarauli vajrauli Ta :
yeṣāmarāli vajrauli Ad :
eṣāmarolī vajrolī Ed.
tadāśājīvitasya ca ] $\beta$ :
tadā meni mameti ca Ba :
tadā mati mateti ca Ed :
mameti ca mameti ca Ad :
tadā meti mameti mateti ca Gb (\textit{unmetr.}) :
mateti mame++ Ta.
}\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_000_13]
The second line of this verse is problematic in both the source text (the \emph{Amaraugha}) and the manuscripts of the \emph{Haṭhapradīpikā}. It seems that the \textepsilon\ group has preserved the most plausible version of this line, which is also quite close to the source text. The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%??JM: comment on pāda d? It's odd, worth cruxing even.
%C2 (Grp 7) tadāmarolī ... sahajolī prajāyate (=Jyotsnā)
\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*14}
\begin{translation}[hp04_000_14]
How can knowledge exist, when the mind is alive, O goddess? So long as the breath lives, the mind does not die. Breath and mind: he who leads this pair to dissolution attains liberation; in no way [does any] other man.
\end{translation}

@@

%\begin{philcomm}[hp04_000_14]
%\end{philcomm}
%?? MD: adopt prayāti (beta) in Pāda C?

@@

%%%%%%%%%%
\subsection*{4.1}
\begin{translation}[hp04_001]
%Now, \emph{samādhi}–
Mind is the master of the senses; breath is the master of the mind [and] dissolution [of the mind] is the master of the breath. [The yogi] should take refuge in that master, dissolution [of the mind].
\end{translation}
%?? JB: Both \alphaOne and \alphaTwo have atha samādhiḥ as a heading before this verse. Shouldnt we include this in our edition or at least note it in the apparatus? We've added atha headings elsewhere from alpha. There's also the issue of where to place atha samādhi in relation to the greyscale verses.

@@

%%%%%%%%%%
\subsection*{4.1*1}
\begin{translation}[hp04_001_1]
Whether or not this [dissolution] is called liberation  in another teaching, there is no difference in the bliss arising from the dissolution of mind and breath.
\end{translation}
%?? JB: 'in another doctrine' for matāntare?
%?? JB: extraordinary bliss (kaścid?)
%\begin{sources}[hp04_001_1]
%\end{sources}

@@

\begin{philcomm}[hp04_001_1]
Some of the \textbeta, \textepsilon\ and \texteta\ groups have a different version of the second half of this verse, the most plausibe of which is, 'the extraordinary bliss of dissolution of mind and breath arises in me' (\emph{manaḥprāṇalayānando mayi kaścit pravartate}). The first person pronoun suggests that this verse derives from a dialogue of some kind before it was changed to the version in the \textgamma\ and \textdelta\ manuscripts.  
\end{philcomm}

% ?? MD: comment on the variant reading "manaḥprāṇalayānando mayi kaścit pravartate"?
% JB: maybe this is the lectio difficilor and the first person pronoun was later written out.

%%%%%%%%%%
\subsection*{4.2}
\begin{translation}[hp04_002]
A motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme for yogis.%??JM: or "The yogis’ motionless and unchanging dissolution... is the best."
\end{translation}
% on the translation of jayati as an intransitive verb, see Alexis' footnote on jayati in his forthcoming book on the Tantrāloka āhnika 1

\begin{sources}[hp04_002]
\emph{Amanaska} 2.21
\begin{versinnote}
\tl{pranaṣṭocchvāsaniśvāsaḥ pradhvastaviṣayagrahaḥ |\\+}
\tl{niśceṣṭo nirgatārambho hy ānandaṃ yāti yogavit ||\\+}
\tl{\var{pranaṣṭocchvāsa°  ] Jb Pa Va Bl Ja Ad AllSI  AllN, 
pranaṣṭasvāsa°  Mb,
pranaṣṭāsvāsa° Ua,
praṇaśocchvāsa° Ma,
pranaṣṭaḥ svāsa° Je Vb,
pranaṣṭo śvāsa°ḥ Pc,
prāṇastho śvāsa° Pb,
pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.??jM: simplify? It's only the ucch that's different. JB: done.
%\tl{\var{hy ānandaṃ yāti yogavit ]  AllNI (except  Ma Pa Pb Pc Tr Va),
%hy ānandaṃ yāti tattvavit Pa Tr Va Ea,
%svānandaṃ yāti yogavit Vd,
%hy ānandayati yoginaḥ Ma Tha,
%hy ānanday[a]ti yogī kaḥ Pc,
%hy ānanda iva yoginaḥ AllN (except  Na Nm Ea Ve),
%hy ānanda iva yogi Na Nm,
%hy ānandaś ca yoginaḥ Ve,
%hy ānandasukhayogavit Pb,
%layo jayati yoginaḥ Ad,
%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.??JM: but it's in almost all HP witnesses too so perhaps variants not needed here, just mention it in comm? Or simplify? JB done
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_002]
%Adopt alpha reading. [MD: Done.]
The fourth verse quarter of this verse differs from that of the North Indian recension of the \emph{Amanaska} (quoted as the source) but is found in two South Indian manuscripts of the \emph{Amanaska} (Ad and Cc).
\end{philcomm}

%%%%%%%%%%
\subsection*{4.3}
\begin{translation}[hp04_003]
May an extraordinary absorption prevail, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%??JM: which is intelligible > intelligible, and other tweaks. Prevail not great but can't think of anything else. 
%% My translation in the Amanaska (reading jāyate): [This] extraordinary absorption by which all volition has been cut off and in which all movement has ceased, is intelligible [only] by means of itself and is beyond the scope of words.
% According to the Amanaska's transmission, jayatāṃ (and its similar readings) arose at a later time.  
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.4}
\begin{translation}[hp04_004]
Dissolution, which is eternal in the elements and senses, takes place where the gaze is. The gaze that has dissolved into the focus becomes the power of living beings.%%??JM: changed meditative focal point to focus (as tr. in v.6)
%JB 'occurs' for 'takes place'
\end{translation}

@@

\emph{Kaulajñānanirṇaya} 3.2
\begin{versinnote}
\tl{yatra dṛṣṭir manas tatra bhūtendriya sapudgalaḥ \\+} 
\tl{svaśaktijīvabhūtā hi dṛṣṭilakṣair layaṃ gatāḥ \\+}
\tl{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed\\!} 
% ?? Consult Shaman's critical edition.
\end{versinnote}
\end{sources}

@@

%%%%%%%%%%
\subsection*{4.6}
\begin{translation}[hp04_006]
The focus internal, the gaze external, unblinking: this is the \emph{śāmbhavī mudrā} concealed in all the Tantras.%??JM: tr. tweaked
% JB my preference would be: The focus is internal [yet] the gaze is external and free from blinking.  
\end{translation}

\begin{sources}[hp04_006]
\emph{Amanaska} 2.10
\begin{versinnote}
\tl{antar lakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitā |\\+}
\tl{eṣā hi śāmbhavī mudrā sarvatantreṣu gopitā ||\\+}
\tl{\var{antar lakṣyaṃ ] AllNI (except Ja K Pb Tr Ua Vb) Tha Vd, antarlakṣya Ad Cc K: antarlakṣaṃ Tr Nw Ve: antarlakṣā Ja Ua Ea Eb, antarlakṣo Na Nq Nu, antarlakṣyo Nb Nm: antaryogaṃ Vb: antarlīno Pb}\\!}%?? JM: delete variants?
% JB: the lakṣyaṃ/lakṣa spelling is an issue in both the Amanaska and HP mss (inlcuding alpha).
\end{versinnote}

@@

\begin{philcomm}[hp04_006]
See Birch 2013: 287 for more parallel verses.
\end{philcomm}

@@

\begin{philcomm}[hp04_008]
This line may be authorial because it is explaining the fact that 4.6 and 4.7 teach similar techniques called \emph{śāmbhavī mudrā} and \emph{khecarīmudrā} respectively. One would expect a verse on \emph{khecarī mudrā} to be explaining the insertion of the tongue in the nasopharyngeal cavity (cf.~3.33–48), but the practice of meditating by fixing the gaze is called \emph{khecarī mudrā} in 4.7 (as attested by \textalpha, \textbeta\ and \texteta\ manuscripts) and manuscripts of the \emph{Candrāvalokana}, the text from which Svātmārāma borrowed this verse. It is therefore likely that he added 4.8 to explain that \emph{śāmbhavī} and \emph{khecarī mudrās} are the same with respect to the gaze.

%Judit would greyscale 4.8, thinking that it is a marginal note. And greyscale the pātāle verses.
%avasthā: Brahmānanda explains this as referring to the place (dhāman) of khecarī (the middle of the brow) and śāmbhavī (the heart).
\end{philcomm}

@@

\begin{sources}[hp04_008_1]
\emph{Amanaska} 1.8 (South Indian Recension)
\begin{versinnote}
\tl{netre jyotiṣi saṃyojya kiṃ cid unnamayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkārakaḥ kṣaṇāt ||\\+}
\tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!}%?? variants needed? JB I've deleted the variants for the last pāda but we need those on unnamayed (see below) 
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_008_1]
%MD: Better to adopt unnamayed and kārakaḥ kṣaṇāt?
% ?? JB Since unnamayed in the beta witnesses and source, and makes better sense, I think we should adopt. The dictionary states that uc-cal means 'to go or move away from; to free or loosen one's self from', which is not so good here.

This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7. 
\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*2}
\begin{translation}[hp04_008_2]
Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over.%??JM: put Tāraka and explain in note? Tārakayoga is mentioned in the previous note.
% JB in the source 
\end{translation}
% Maybe grey scale

@@

\begin{philcomm}[hp04_008_2]
On why this verse is in greyscale, see the note to 4.8*1.
% ??JB we should adopt nigama in pāda b because it is attested by several manuscript groups and the source.
\end{philcomm}

@@

\begin{sources}[hp04_008_3]
\emph{Candrāvalokana} 3
\begin{versinnote}
\tl{ardhotghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
\tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
\tl{tatvaṃ tatpadam eti yat tu paramaṃ vācyaṃ kim atrādhikaṃ ||\\+}
\tl{\var{sthiramanā ] 4340 4345 T00788, sphuramanā 7970}\\+}
\tl{\var{nāsāgradattekṣaṇaś ] 7970 4345 T00788, nāsāgradaṭhattakṣaṇaś 4340}\\+}
\tl{\var{tatvaṃ tatpadam eti yat tu] 4340 4345 T00788, tatvaṃ-n-tatpadam eti yat kṛ 7970}\\!}
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_008_3]
This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze. In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups.
\end{philcomm}
% ?? MD: I have changed the verse order according to the epsilon group. Please adjust the commentary. JB Done

@@

%\begin{philcomm}[hp04_008_7]
%?? JM: nirodhataḥ reading much better. Early corruption?
% JB I think we should note nirodhataḥ as the better reading in the comments but adopt divārātrau (or divārātraṃ) na pūjayet (he should not worship at day and night (i.e., when prāṇa is moving in the sun and mmoon channels)). 
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*8}
\begin{translation}[hp04_008_8]
Now \emph{khecarī}:\\
There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
\end{translation}
% ?? compound khecarīmudrā in 4.8*8c and 4.8*9c?

%\begin{sources}[hp04_008_8]
%\end{sources}

@@

\begin{testimonia}[hp04_008_9]
\emph{Haṭhasaṅketacandrikā} f.~??
\begin{versinnote}
\tl{savyadakṣiṇanāḍīstho madhye carati mārutaḥ |\\+}
\tl{tiṣṭhati khecarī mudrā tasmin sthāne na saṃśayaḥ ||\\!}
\end{versinnote}
\end{testimonia}

@@

\subsection*{4.8*16}
\begin{translation}[hp04_008_16]
Between the eyebrows is the place of Śiva. The mind dissolves there. That level should be known as the fourth state. Death does not exist there.
\end{translation}
%?? JB: pada here means place/site (level sounds strange)
% abode of Śiva ... that place should be known

@@

\emph{Upāsanāsārasaṅgraha} p.~106
\begin{versinnote}
\tl{audāsīnyaparo bhūtvā sadābhyāsena samyamī |\\+}
\tl{unmanīkaraṇam sadyo nādam evānu [...] yet ||\\!}
\end{versinnote}
\end{testimonia}

>>>>

%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY. Judit: kalā is like śakti (initial source)?
%Refer to TĀK

@@

%%%%%%%%%%
\subsection*{4.0*7}
\begin{translation}[hp04_000_7]
Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation is established [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
\end{translation}

@@

\begin{philcomm}[hp04_000_7]
The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}. Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}. The compound \emph{muktisthitiḥ} (literally, `the establishment of liberation') has been adopted. 
%  attested by \etaTwo, 
% adopt muktisthitiḥ and rewrite note. 
%
\end{philcomm}

@@

\begin{philcomm}[hp04_000_9]
In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of Suṣumnā.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.0*10}
\begin{translation}[hp04_000_10]
For the yogi whose Kuṇḍalinī has awakened and who has given up all activity, the innate state automatically shines forth.
\end{translation}

@@

%

%%%%%%%%%%
\subsection*{4.0*11}
\begin{translation}[hp04_000_11]
When the breath is flowing in the central channel, and the mind enters the void, the knower of yoga destroys all actions.
\end{translation}
%?

@@

%\begin{philcomm}[hp04_000_12]
%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
%

@@

%%%%%%%%%%
\subsection*{4.0*13}
\begin{translation}[hp04_000_13]
When equanimity has been obtained, and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises †even for the person whose hope is for that†.
\end{translation}
%

\begin{sources}[hp04_000_13]
\emph{Amaraugha} 7
\begin{versinnote}
\tl{citte tu sattvam āpanne vāyau vrajati madhyame |\\+}
\tl{tadāmaraughavajrolis tadāśājīvitasya ca ||\\+}
\tl{\var{citte tu sattvam ] Ga, Ae (=\emph{Amaraugha}) : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
tadāmaraughavajrolis ] \textit{conj.} :
tadāmaraughavajroli Ga :
tadāmaraudhavajrolī Ae :
eṣāmarauli vajroli Gb :
eṣāmarāli vajroli Ba :
eṣāmarauli vajrauli Ta :
yeṣāmarāli vajrauli Ad :
eṣāmarolī vajrolī Ed.
tadāśājīvitasya ca ] Ga, Ae (=\emph{Amaraugha}) :
tadā meni mameti ca Ba :
tadā mati mateti ca Ed :
mameti ca mameti ca Ad :
tadā meti mameti mateti ca Gb (\textit{unmetr.}) :
mateti mame++ Ta.
}\\!}
\end{versinnote}

@@

\begin{philcomm}[hp04_000_13]
The second line of this verse is problematic in both the source text (the \emph{Amaraugha}) and the manuscripts of the \emph{Haṭhapradīpikā}. For the last \emph{pāda}, we have adopted the closest reading to that of the source text (the \emph{Amaraugha}) but cannot make sense of it so have cruxed it. The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%
%C2 (Grp 7) tadāmarolī ... sahajolī prajāyate (=Jyotsnā)
\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*14}
\begin{translation}[hp04_000_14]
How can knowledge exist, when the mind is alive, O goddess? So long as the breath lives, the mind does not die. Breath and mind: [when] this pair goes to dissolution, a man attains liberation; in no way [does any] other [man].
\end{translation}

@@

%\begin{philcomm}[hp04_000_14]
%\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.1}
\begin{translation}[hp04_001]
%Now, \emph{samādhi}–
Mind is the master of the senses; breath is the master of the mind [and] dissolution [of the mind] is the master of the breath. [The yogi] should take refuge in that master, dissolution [of the mind].
\end{translation}
%

@@

%%%%%%%%%%
\subsection*{4.1*1}
\begin{translation}[hp04_001_1]
Whether or not this [dissolution] is called liberation in another school, an extraordinary bliss from the dissolution of mind and breath arises in me.
\end{translation}
%\begin{sources}[hp04_001_1]
%\end{sources}

@@

\begin{philcomm}[hp04_001_1]
The manuscripts attest two versions of the second line. It appears that the reading of \textbeta, \textepsilon\ and \texteta\ groups, which we have adopted, was changed to remove the first person pronoun. The modified reading is prevalent in the \textgamma\ and \textdelta\ manuscripts (\emph{manaḥprāṇalayānando nāpi kaś cit vibhidyate}).  
\end{philcomm}
%

%%%%%%%%%%
\subsection*{4.2}
\begin{translation}[hp04_002]
The yogis' motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme.
\end{translation}

\begin{sources}[hp04_002]
\emph{Amanaska} 2.21
\begin{versinnote}
\tl{pranaṣṭocchvāsaniśvāsaḥ pradhvastaviṣayagrahaḥ |\\+}
\tl{niśceṣṭo nirgatārambho hy ānandaṃ yāti yogavit ||\\+}
\tl{\var{pranaṣṭocchvāsa° ] Jb Pa Va Bl Ja Ad AllSI  AllN, 
pranaṣṭasvāsa°  Mb,
pranaṣṭāsvāsa° Ua,
praṇaśocchvāsa° Ma,
pranaṣṭaḥ svāsa° Je Vb,
pranaṣṭo śvāsa°ḥ Pc,
prāṇastho śvāsa° Pb,
pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.
%\tl{\var{hy ānandaṃ yāti yogavit ]  AllNI (except  Ma Pa Pb Pc Tr Va),
%hy ānandaṃ yāti tattvavit Pa Tr Va Ea,
%svānandaṃ yāti yogavit Vd,
%hy ānandayati yoginaḥ Ma Tha,
%hy ānanday[a]ti yogī kaḥ Pc,
%hy ānanda iva yoginaḥ AllN (except  Na Nm Ea Ve),
%hy ānanda iva yogi Na Nm,
%hy ānandaś ca yoginaḥ Ve,
%hy ānandasukhayogavit Pb,
%layo jayati yoginaḥ Ad,
%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_002]
The fourth verse quarter of this verse differs from that of the North Indian recension of the \emph{Amanaska} (quoted as the source) but is found in two South Indian manuscripts of the \emph{Amanaska} (Ad and Cc).
\end{philcomm}

%%%%%%%%%%
\subsection*{4.3}
\begin{translation}[hp04_003]
May an extraordinary absorption reign supreme, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%
%% My translation in the Amanaska (reading jāyate): [This] extraordinary absorption by which all volition has been cut off and in which all movement has ceased, is intelligible [only] by means of itself and is beyond the scope of words.
% According to the Amanaska's transmission, jayatāṃ (and its similar readings) arose at a later time.  
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.4}
\begin{translation}[hp04_004]
Dissolution, which is eternal in the elements and senses, occurs where the gaze is. The gaze that has dissolved into the focus becomes the power of living beings.%%
%
\end{translation}

@@

\emph{Kaulajñānanirṇaya} 3.2
\begin{versinnote}
\tl{yatra dṛṣṭir manas tatra bhūtendriya sapudgalaḥ \\+} 
\tl{svaśaktijīvabhūtā hi dṛṣṭilakṣair layaṃ gatāḥ \\+}
\tl{\var{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed}\\!} 
% 
\end{versinnote}
\end{sources}

@@

%%%%%%%%%%
\subsection*{4.6}
\begin{translation}[hp04_006]
The focus is internal, the gaze external, unblinking: this is the \emph{śāmbhavī mudrā} concealed in all the Tantras.%

\begin{sources}[hp04_006]
\emph{Amanaska} 2.10
\begin{versinnote}
\tl{antar lakṣyaṃ bahir dṛṣṭir nimeṣonmeṣavarjitā |\\+}
\tl{eṣā hi śāmbhavī mudrā sarvatantreṣu gopitā ||\\+}
\tl{\var{antar lakṣyaṃ ] AllNI (except Ja K Pb Tr Ua Vb) Tha Vd, antarlakṣya Ad Cc K: antarlakṣaṃ Tr Nw Ve: antarlakṣā Ja Ua Ea Eb, antarlakṣo Na Nq Nu, antarlakṣyo Nb Nm: antaryogaṃ Vb: antarlīno Pb}\\!}%JM: delete variants?
% JB: the lakṣyaṃ/lakṣa spelling is an issue in both the Amanaska and HP mss (inlcuding alpha).
\end{versinnote}

@@

\begin{philcomm}[hp04_006]
As seen in the witnesses of the \emph{Amanaska}, the source text, the spellings \emph{lakṣya} and \emph{lakṣa} occur randomly in the manuscripts. See Birch 2013: 287 for more parallel verses.
\end{philcomm}

@@

\begin{philcomm}[hp04_008]
This line may be authorial because it is explaining the fact that 4.6 and 4.7 teach similar techniques called \emph{śāmbhavī mudrā} and \emph{khecarīmudrā} respectively. One would expect a verse on \emph{khecarī mudrā} to be explaining the insertion of the tongue in the nasopharyngeal cavity (cf.~3.33–48), but the practice of meditating by fixing the gaze is called \emph{khecarī mudrā} in 4.7 (as attested by \textalpha, \textbeta\ and \texteta\ manuscripts) and manuscripts of the \emph{Candrāvalokana}, the text from which Svātmārāma borrowed this verse. It is, therefore, likely that he added 4.8 to explain that \emph{śāmbhavī} and \emph{khecarī mudrās} are the same with respect to the gaze. A similar conflation occurs in the \textit{Śivayogapradīpikā} (5.3).
% NJ: Maybe mention Śivayogapradīpikā 5.3 and 5.4 
%% 5.3ab lakṣyalīnamanasānilena yo vartate 'calitatārako bhavet |
%% 5.3cd khecarīyam atha saiva śāṃbhavīmudrayā sakṛtayā jagadguruḥ 
%% (5.3) For one whose eyes remain unmoving [while open], with the mind and breath dissolved in the [inner] lakṣya, this khecarī indeed is śāmbhavī. Through the proper practice of this mudrā, he shall be the guru of the world.
%% 5.4ab loke 'antaḥkhecarī mudrām ajñātvā bāhyakhecarīm
%% 5.4cd avalambya tvarāḥ kecij jihvāchedanakarmaṇā
%% (5.4) Not knowing the inner khecarīmudrā in this world, some have hastily taken to the external khecarī through the practice of cutting the tongue.
%%%%
%Judit would greyscale 4.8, thinking that it is a marginal note. And greyscale the pātāle verses.
%avasthā: Brahmānanda explains this as referring to the place (dhāman) of khecarī (the middle of the brow) and śāmbhavī (the heart).
\end{philcomm}

@@

\begin{sources}[hp04_008_1]
\emph{Amanaska} 1.8 (South Indian Recension)
\begin{versinnote}
\tl{netre jyotiṣi saṃyojya kiṃ cid unnamayed bhruvau |\\+}
\tl{pūrvayogasya mārgo 'yam unmanīkārakaḥ kṣaṇāt ||\\+}
\tl{\var{unnamayed ] Cb Vd: unnamayan Bb Tha: unmīlaye W: unmilayet Uc: unmanaya Pe: unmīlya Cc}\\!}
%\tl{\var{unmanīkārakaḥ kṣaṇāt ] Bb Tha W Vd Pe: unmunikārakakṣaṇāt Uc: unmanīkāraṇaṃ kṣaṇāt Cc: unmanākāśakaḥ kṣaṇāt Ad Ca: unmanākāśakakarṣaṇāt Cb}\\!} 
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_008_1]
This and the next verse (4.8*1–2) are from the South Indian recension of the \emph{Amanaska}, a later rewriting of that work which incorporates these verses in an additional passage on Tārakayoga, added here to elaborate on the meditation of fixing the gaze that is taught in 4.4–7. 
\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*2}
\begin{translation}[hp04_008_2]
Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
\end{translation}

@@

\begin{philcomm}[hp04_008_2]
In the source text \emph{tārakam} refers to  Tāraka yoga, one of two yogas taught in the South Indian recension of the \emph{Amanaska}.

On why this verse is in greyscale, see the note to 4.8*1.    
\end{philcomm}

@@

\begin{sources}[hp04_008_3]
\emph{Candrāvalokana} 3
\begin{versinnote}
\tl{ardhodghāṭitalocana[ḥ] sthiramanā nāsāgradattekṣaṇaś\\+}
\tl{candrārkāv api līnatām upagatau niṣpandarūpaṃ vapuḥ |\\+}
\tl{jyotīrūpam aśeṣabāhyarahitaṃ dedīpyamānaṃ paraṃ\\+}
\tl{tatvaṃ tatpadam eti yat tu paramaṃ vācyaṃ kim atrādhikaṃ ||\\+}
\tl{\var{sthiramanā ] 4340 4345 T00788, sphuramanā 7970}\\+}
\tl{\var{nāsāgradattekṣaṇaś ] 7970 4345 T00788, nāsāgradaṭhattakṣaṇaś 4340}\\+}
\tl{\var{tatvaṃ tatpadam eti yat tu] 4340 4345 T00788, tatvaṃ-n-tatpadam eti yat kṛ 7970}\\!}
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp04_008_3]
This verse is not in the \textalpha\ manuscripts and was probably added because it elaborates on the gaze. In the \emph{Candrāvalokana} (i.e., the source text) and the \emph{Anubhavanivedana}, it follows 4.7, which may explain its position in the \textgamma\ and \textdelta\ groups.
\end{philcomm}

@@

\begin{philcomm}[hp04_008_7]
The manuscripts transmit readings for the last verse quarter that either contradict or repeat the statements in the first line. This problem likely occurred through some kind of dittographical error. We have therefore adopted the reading of \emph{Jyotsnā} 4.42d (°nirodhataḥ), which is the same as the source text. 
\end{philcomm}

%%%%%%%%%%
\subsection*{4.8*8}
\begin{translation}[hp04_008_8]
Now \emph{khecarī}:\\
There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
\end{translation}
%
%\begin{sources}[hp04_008_8]
%\end{sources}

@@

%\begin{testimonia}[hp04_008_9]
%\emph{Haṭhasaṅketacandrikā} f. 246v (Mysore ORI Manuscript B220) 
%\begin{versinnote}
%\tl{savyadakṣiṇanāḍīstho madhye carati mārutaḥ |\\+}
%\tl{tiṣṭhati khecarī mudrā tasmin sthāne na saṃśayaḥ ||\\!}
%\end{versinnote}
%\end{testimonia}

@@

\subsection*{4.8*16}
\begin{translation}[hp04_008_16]
Between the eyebrows is the place of Śiva. The mind dissolves there. That level should be known as the fourth state. Death does not exist there.
\end{translation}
%

@@
